---
title: "Data Wrangling in R"
author: "Patrick Chester at NYU Data Services: data.services@nyu.edu"
date: '`r format(Sys.time(), "%d %B, %Y")`'
output: html_document
---

Last edited: 8/22/2019

# I. Brief Overview of Data Types

Remove all objects and packages in your current environment

```{r}
rm(list = ls())
```

##  Vectors 

```{r}
v1 <- rnorm(100, 75, 15)
v2 <- as.factor(rep(c("A", "B", "C", "D"), times = 25))
v3 <- rnorm(100, 1, .5) 
```

You can subset vectors using index numbers
```{r}
v2[1:10]
v2[c(4, 8, 12)]
```

## Data Frames 

Extracting variables from a data frame using a \$ operator.
```{r}
mydata <- data.frame(v1, v2, v3)

mydata$v1
```

Selecting a subset of a data frame using row indexes and column names
```{r}
mydata[1:10, c("v1", "v2")]
```

Other methods of data type conversion:
- as.numeric() 
- as.character()
- as.vector() 
- as.matrix() 
- as.data.frame()
- as.factor()


# II. Data Management

## A. Working with Strings

### Regular expressions

*grep:* Identifying strings that match a particular criteria
```{r}
fruit <- c('Apple', 'Banana', 'Orange', 'Grape', 'Pineapple', 'Kiwi', 'Peach', 'Mango', 'Strawberry', 'Guava', 'Cherry')

grep(pattern = 'a', fruit, value = TRUE, ignore.case = TRUE)
```

*gsub:* Replacing instances of a string with another of your choosing
```{r}
gsub(pattern = "Ch", replacement = "B", fruit[11])
```

*gregexpr:* Identifying positions of a string of interest
```{r}
(positions_a <- gregexpr(pattern = "a", text = fruit, ignore.case = TRUE))
```


### Paste

Use seperators to combine strings with other characters or numbers
```{r}
paste("X", 1:5, sep = ".")
```

Use collapse to combine multiple string outputs together
```{r}
paste("X", 1:5, sep = ".", collapse = "")
```

paste0 is the same as paste with an empty seperator
```{r}
paste0("X", 1:5)
```

NOTE: paste0 does not have a "sep" option that can be modified by the user

paste0("a", "b", sep = "c") == paste0("a", "b", "c")

paste0(rep(c("A","C","G","T"), each=4), c("A","C","G","T"), collapse = "")

### Extra material: Other string manipulations

Changing the case of strings
```{r}
string1 <- 'NYU Data Services'
tolower(string1) 
toupper(string1)
```

### Exercise 1: Exercise with Strings

Create this string 'A&1B&2C&3' using a paste function



## B. Working with Dates

```{r}
dates <- c('11/28/2011', '12/07/2012', '08/01/2013', '02/09/2015')
class(dates)
```

```{r}
real_dates <- as.Date(dates, format = '%m/%d/%Y')
class(real_dates)
```

```{r}
other_format <- format(real_dates, '%A %B %d, %Y')
class(other_format)
```


For the codes used to identify and format dates:
```{r}
?strptime
```

Identifying how long ago dates occurred
```{r}
today <- Sys.Date()

(dif <- today - real_dates)

class(dif)
```

To make a difference in times with a particular time unit of interest use difftime

```{r}
difftime(today, real_dates, units = "hours")
```


### Extra Materials: Lubridate

The lubridate package contains a powerful set of tools that can be used to extract and interact with dates.

There are functions like *mdy* that allow for simpler extration of date information from strings.
```{r}
#install.packages("lubridate")
library(lubridate, quietly = TRUE)

(lubri_dates <- mdy(dates))
```

Extracting specific date information from a date object
```{r}
year(lubri_dates)

month(lubri_dates)

day(lubri_dates)
```


### Exercise 2: Exercise with Dates

a. Take the following date (November 11, 2011) and turn it into a date vector in R
b. Display the date vector in the format (month.day.year')

## C. Subsetting Data Sets


Loading UScereal data into our environment from the MASS package
```{r}
library(MASS)
data(UScereal)
head(UScereal)
```

### Selecting Cases using base R

Identifying index numbers of interest
```{r}
which(UScereal$mfr == 'K')  # reports indicies       
which(UScereal$calories > 250)
```

Use indices to select rows w/ selected columns                                              
```{r}
UScereal[which(UScereal$mfr == 'K'), c('mfr', 'calories')]
```

*subset:* subset data frames in base R 
```{r}
subset(UScereal, calories > 250, c('mfr', 'calories'))
```


### Select rows and columns with dplyr, using the %>% operator

*filter:* select rows
```{r}
# install.packages('dplyr')

library(dplyr, quietly = TRUE)

UScereal %>% filter(calories > 250)
```

*select:* Select columns 
```{r}
UScereal %>% select(mfr, calories)
```

Select rows and columns by combining *filter* and *select* with *%>%*
```{r}
UScereal %>% filter(calories > 250) %>% select(mfr, calories)
```

## D. Sorting Data 

### Sorting vectors

*sort:* the sort function only works for vectors
```{r, error=TRUE}
sort(UScereal$calories)  
sort(UScereal)
```

### Sorting data frames using index numbers

Order produces index numbers of input vector arranged so that output vector is arranged in ascending order

```{r}
order(UScereal$calories)  

UScereal[order(UScereal$calories), c('mfr', 'calories')] 

UScereal[order(UScereal$mfr, -UScereal$calories), ]
```


### Sorting with dplyr

```{r}
library(dplyr, quietly = TRUE)

UScereal %>% arrange(mfr, desc(calories)) 
```



### Exercise 3: Exercise with subsetting

Create a subset of mydata, which contains the 25 highest v1 scores


## E. Converting panel data from long to wide and vice versa 

Make a Panel Dataset
```{r}
health <- data.frame(id = rep(1:10, each = 4, len = 40),
                     trial = rep(c(1:4), 10), 
                     score = rnorm(40, 3, .75))

head(health, n = 5)
```

### Reshaping with the reshape function

Long --> Wide
```{r}
health_wide <- reshape(health, v.names = "score", idvar = "id", 
                       timevar = "trial", direction = "wide") 

tail(health_wide, n = 5)
```


Wide --> Long
```{r}
health_long <- reshape(health_wide, idvar = "id", direction = "long")         

head(health_long, 5)
```

### Reshaping panel data using the tidyr package

*spread:* Convert long data to wide
```{r}
# install.packages('tidyr')
library(tidyr, quietly = TRUE)

spread(health, key = trial, value = score) # key is the identifier
```

*gather:* Convert wide data to long
```{r}
gather(health_wide, key = trial, value = score, score.1:score.4) # can also reference by column number(2:5)
```

### Exercise 4: Exercise with reshaping

Determine if the dataset below is long or wide, and reshape 

```{r, eval=FALSE}
data_reshape <- data.frame(State = rep(state.name, 4), year = rep(2010:2013, each = 50), GSPPC = rnorm(200, mean = 40000, sd = 10000))
```


## F. Merge Datasets 

Create two data with common variables
```{r}
(data1 <- data.frame(id = rep(1:5, 3), year = rep(2000:2002, each = 5), 
                     group = sample(c("A", "B", "C"), 15, replace = TRUE)))

```

```{r}
(data2 <- data.frame(id = rep(1:5, each = 4), year = rep(2000:2003, 5),
                    score = rnorm(20, 50, 15)))
```

### Merging using base R's merge function

Merge them by id & year, 1:1 merge
```{r}
data_merge <- merge(data1, data2, by = c("id", "year")) 

head(data_merge)
```

Is there anything odd about the merge? *Hint:* look at the year.


Extra rows from both datasets are added.

```{r}
data_merge <- merge(data1, data2, by = c("id", "year"), all = TRUE) 

head(data_merge)
```

If you wanted to do a left merge: x.all=TRUE
A right merge: y.all=TRUE

### Merging with the dplyr package

*inner_join*
```{r}
library(dplyr, quietly = TRUE)

inner_merge <- data1 %>% inner_join(data2, by = c("id", "year"))

head(inner_merge)
```

*full_join*
```{r}
outer_merge <- data1 %>% full_join(data2, by = c("id", "year"))

head(outer_merge)
```


### Extra Materials: Data Services' merging documentation for R and other statistical software
```{r}
browseURL("http://guides.nyu.edu/quant/merge")
```

### Exercise 5: Exercise with merging

Merge the following data set to data1

```{r}
(data3 <- data.frame(id = rep(1:5, each = 4), year = rep(2000:2003, 5),
                    response = sample(1:5, size = 20, replace = TRUE)))
```


## G. Apply function and Aggregate Statistics 

### Apply functions
Apply mean() across numeric nutrition variables in the UScereal dataset     

With MARGIN = 1, apply calculates the *mean* over rows.
```{r}
apply(UScereal[, 2:9], MARGIN = 1, FUN = mean)  
```

With MARGIN = 2, apply calculates the *mean* over columns.
```{r}
apply(UScereal[, 2:9], MARGIN = 2, FUN = mean)
```

Apply sd() across numeric nutrition variables in the UScereal dataset      
```{r}
lapply(UScereal[, 2:9], sd)
sapply(UScereal[, 2:9], sd)
```


An application of sapply to our vector of fruit names from section 1
```{r}
(num_a <- sapply(positions_a, function(x) ifelse(x[1] > 0, length(x), 0)))
```

### Aggregation with Base R 

*tapply:* Summarizes a variable by a grouping variable and return a vector or an array.
```{r}
tapply(UScereal$calories, UScereal$mfr, summary)
```

*aggregate:* Similar to tapply, but it outputs a data frame.
```{r}
aggregate(UScereal$calories, list(UScereal$mfr), mean)
```


The by function can also provide Summary statistics of 'calories' grouped by 'mfr'            
by(UScereal\$calories, UScereal\$mfr, summary)
by(UScereal\$calories, list(UScereal\$mfr, UScereal\$shelf), summary) 

### Aggregation with dplyr

Use *group_by* and *summarize* to produce a data frame that produces summary statistics over your grouping variable(s)

```{r}
library(dplyr, quietly = TRUE)

UScereal %>% 
  group_by(mfr) %>% 
  summarize(avg.cal = mean(calories))
```

You can compute multiple summary statistics at once, such as a count of the number of cereals produced by each manufacturer.
```{r}
UScereal %>% 
  group_by(mfr) %>% 
  summarize(avg.cal = mean(calories), count = n())
```

dplyr's *mutate* function allows users to create variables. When combined with *group_by* it allows for the creation of summary variables that are added to the original data frame.
```{r}
UScereal %>% 
  group_by(mfr) %>% 
  mutate(avg.cal = mean(calories), count = n())
```


### Exercise 6: Exercise with aggregation

Using one of the methods above, find the median protein of cereals by each shelf 



# III. Functions

Structure of a function in R
```{r, eval=FALSE}
function.name <- function(arg1, arg2, ...) {
  function_body
  return(any_value_to_return)
  }
```



Create a simple function that adds two numbers: 'a' and 'b'
```{r}
addTwoNums <- function(a, b) {
  tmp <- a + b
  return(tmp)
  # Alternatively either of the below would substitute for the above
  # return(a + b)
  # a + b
}

addTwoNums(2, 1)
```

Does this work? Why or why not?
```{r, error=TRUE}
addTwoNums(5)
```

We can set default values for parameters in a function
```{r}
addTwoNums <- function(a, b = 2) {
  return(a + b)
}

addTwoNums(5)
```

Simple functions are vectorized in R by default, allowing you to run your function over the elements of a vector (assuming other inputs are constant).
```{r}
addTwoNums(a = c(4, 10, 0))
```

What happens in this case?
```{r}
addTwoNums(3, 4)
```


Multiple results to report 

```{r}
myOperations <- function(a, b) {
  add <- a + b
  subtract <- a - b
  multiply <- a * b
  divide <- a / b
  mylist <- list(add, subtract, multiply, divide)
  return(mylist)
}

myOperations(5, 10)
```


## Extra Materials

Graph a function 
```{r}
f <- function(x) (x ^ 3 - 13)

plot(-5:6, #x values
     f(-5:6), #y values
     type = "l", #line plot
     ylab = "f(x) = x ^ 3 - 13", xlab = "") #labels
```

*switch:* A function that returns different output depending on the input
Returns output depending on an index number
```{r}
switch(1, c("one", "two"),c("three", "four"))
```

*switch:* Returns output depending on labels
```{r}
switch("second", first = c("one", "two"), second = c("three", "four"))
```

A function that takes a numeric vector and performs a significance test with user defined parameters.
```{r}
my_ttest <- function(x, mu = 0, test = "two-sided", alpha = 0.05) {
  #calculating values, independent of test type
  n <- length(x)
  df <- n - 1
  std <- sqrt(var(x))
  t <- sqrt(n) * (mean(x) - mu) / std 
  
  #p value dependent on test type
  tail_area <- switch(test,  
                      "two-sided" = 2 * (1 - pt(abs(t), n - 1)),
                      "lower" = pt(abs(t), df), 
                      "upper" = 1 - pt(abs(t), df))
  list(t.statistics = t, degree.freedom = df, p.value = tail_area)
}

my_ttest(v1)

```

### Exercise 7: Exercise with functions

Create a function that identifies the second largest element in a numeric vector

# IV. For Loop

Loops allow you to iterate over inputs to produce outputs in a similar way we saw with the apply functions. In fact, apply functions are themselves loops! Writing your own loops gives you the ability to solve more complex problems that couldn't be easily solved with the suite of apply functions.

Simple loop that prints out university names where *univ* is the vector if inputs and *i* is the variable of the loop.
```{r}
univ <- c("Yale", "Harvard", "Pittsburg", "NYU", "Rutgers")
for (i in univ) {
  print(i)
}
```

A more sophisticated loop where each i th element in the output is the sum of i and all previous values of the input vector.
```{r}
out <- 0
for (i in c(1, 2, 4, 5))  {
  out <- out + i
  print(out)
}
```

Identify the cereals with the highest calories, protein, and fat
```{r}
var_names <- c("calories", "protein", "fat")

# Create a matrix that will contain the names of the top 10 cereals
mat <- matrix(nrow = 10, ncol = 3)

for(i in 1:length(var_names)){
  mat[, i] <- rownames(UScereal[order(UScereal[, var_names[i]]), ])[1:10]
}

#when reading a long line of code like the one in 615, 
#try to understand it by working from the inside out or
#testing the code from the outside in
#inside out: var_names[i] i the i th variable name,...
#code testing: copy line of code line 615 from "rownames" on and
#replace var_names with a variable name like "calories"
#delete different parts of the line and see what changes


colnames(mat) <- var_names
 
(mat <- as.data.frame(mat))
```


### Exercise 8: Exercise with functions

Create a function that produces a data frame that contains the names of the five cereals with the lowest sodium, fiber, and sugars. Hint: see the above example


**END OF LECTURE**

# Evaluation

Please help us improve this tutorial and others by taking the survey below:
Just run the line below to open the evaluation in your browser, or copy this link http://bit.ly/ManagementR

```{r}
browseURL("http://bit.ly/ManagementR")
```

If you are interested in additional resources to learn about R, check out r-bloggers at the link below:

```{r}
browseURL("https://www.r-bloggers.com/")
```

Consider the material below to be supplemental.

# V. If/Else Statements

## If and else statements

If and else statements allow you to specify outputs that depend on the values of an input
```{r}
x <- 10
if (x > 10) {
  print ("Greater than 10")
} else if (x == 10) {
  print ("Equal to 10")
} else if (x < 10 && x >= 0) {
  print ("Between 0 and 10")
} else {
  print ("Less than 0")
}
```


## The ifelse function 

This function allows you to apply a test over a vector where the output for each vector element is defined in the *ifelse* function.
```{r}
x <- seq(1:4)

ifelse(test = x < 4, # The test
       yes = "less than 4", # What the function returns when the test is true
       no = "more than equal to 4") # What is returned when the test is false
```


# VI. While Loop

While loops allow you to specify a condition that will end the loop.

This while loop ends when i is greater than 5.
```{r}
i <- 1
while (i <= 5) {
  i <- i + 2
  print(i)
}
```

Here, the while loop ends when univ takes the value of Rutgers.
```{r}
univ <- c("Yale", "Harvard", "Pittsburg", "NYU", "Rutgers")
i <- 1
while(univ[i] != "Rutgers") {
  print(univ[i])
  i <- i + 1
}
```

Creating a function with a while loop 
```{r}
univ_while <- function(x) {
  i <- 1
  while(x[i] != "NYU") {
    print(x[i])
    i <- i + 1
  }
}

univ_while(univ)

```

# VII. Repeat Loop

Repeats an action until a *break* condition is met. Functions similarly to the *while* loop, except the *break* is located in the function itself.

```{r}
i <- 2
repeat {
  print(i)
  i <- i + 2
  if (i > 20) break
}
```

Making the univ_while function with repeat and break (calling it f-repeat)
```{r}
univ <- c("Yale", "Harvard", "Pittsburg", "NYU", "Rutgers")   
f_repeat <- function(x) {
  i <- 1
  repeat { 
  print(x[i])
  i <- i + 1
  if (x[i] == "NYU") break
  }
}

f_repeat(univ)
```

# Exercise Solutions

Of course there is always more than 1 solution, below are just examples

## Exercise 1: Exercise with Strings

Create this string 'A&1B&2C&3' using a paste function

```{r}
paste(c('A', 'B', 'C'), 1:3, sep = '&', collapse = '')
```

## Exercise 2: Exercise with Dates

a. Take the following date (November 11, 2011) and turn it into a date vector in R
b. Display the date vector in the format (month.day.year')

```{r}
date_ex1 <- 'November 11, 2011'

a1 <- as.Date(date_ex1, '%B %d, %Y')

format(a1, '%m.%d.%y')
```


## Exercise 3: Exercise with subsetting 

Create a subset of mydata, which contains the 25 highest v1 scores

```{r}
sorted <- mydata[order(-mydata$v1), ][1:25, ]

fullsort<-mydata %>% arrange(desc(v1))
fullsort[1:25, ]
```

## Exercise 4: Exercise with reshaping 

Determine if the dataset below is long or wide, and reshape

```{r}
data_reshape <- data.frame(State = rep(state.name, 4), Year = rep(2010:2013, each = 50), GSPPC = rnorm(200, mean = 40000, sd = 10000))

#long data set, so we'll make it wide
reshape(data_reshape, v.names = "GSPPC", idvar = "State",
        timevar = "Year", direction = "wide")
```


## Exercise 5: Exercise with merging

Merge the the new data set with data1

```{r}
merge(data1, data3, by = c("id", "year"))
```

## Exercise 6: Exercise with aggregation

Using one of the methods above, find the median protein of cereals by each shelf

With *aggregate*
```{r}
aggregate(UScereal$protein, by = list(UScereal$shelf), FUN = median)
```

With *group_by* and *summarize*
```{r}
UScereal %>% 
  group_by(shelf) %>% 
  summarise(median(protein))
```

## Exercise 7: Exercise with functions

Create a function that identifies the second largest element in a numeric vector

```{r}
vec <- 1:10

max2 <- function(v) {
  sort(v, decreasing = TRUE)[2]
}

max2(vec)
```

## Exercise 8: Exercise with loops

Create a function that produces a data frame that contains the names of the five cereals with the lowest sodium, fiber, and sugars

```{r}
var_names <- c("sodium", "fibre", "sugars")

# Create a matrix that will contain the names of the top 5 cereals

top_5_cereals <- function(x) {
  
  mat <- matrix(nrow = 10, ncol = 3)

  for(i in 1:length(var_names)){
    mat[, i] <- rownames(UScereal[order(UScereal[, var_names[i]]), ])[1:5]
  }
  
  colnames(mat) <- var_names
   
  (mat <- as.data.frame(mat))
}

top_5_cereals(var_names)
```
